---
title: "Model Planning & Building"
output:
  html_document:
    df_print: paged
---

This builds upon the first part, from [`deliverable1.Rmd`](deliverable1.html)
```{r echo=FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}
include <- function(library_name){
  if( !(library_name %in% installed.packages()) )
    install.packages(library_name) 
  library(library_name, character.only=TRUE)
}
include("rvest")
include("tidyr")
include("tidyverse")
include("ggplot2")
include("modelr")
include("knitr")
library(caret)
purl("deliverable1.Rmd", output = "deliverable1.r") # produces r source from rmd

# Since the first deliverable requires user to download a file (See note below) this part will be commented out - once file is downloaded this line should be uncommented.
#source("deliverable1.r") # executes the source
```
**NOTE**: This deliverable file is modular however the file "uspollution.csv" will need to be downloaded on your local computer before running. This file is too large to be included within the project's repository.

# **Supplemental Data**
## Load Data from Database
We are using a dataset on U.S. greenhouse gas emissions by inventory sector recorded from 1990 to 2017. The five sectors are Energy, Agriculture, Industrial Processes, Waste and Land-use/Change/Forestry. This data is loaded in and exported as a CSV file using the EPA's Database. The data can be explored and downloaded at the [`EPA Database`](https://cfpub.epa.gov/ghgdata/inventoryexplorer/#iallsectors/allgas/inventsect/all)

**NOTE**: The supplemental dataset I used is predownloaded and saved within this project for easier access since multiple datasets can be generated through the database, the dataset provided ("InventorySector.csv") is specific to this project.
```{r}
inventory_sector <- read_csv("InventorySector.csv")
# Priview of dataset
head(inventory_sector)
```

## Tidy Up
Let's clean up the dataset by making each sector it's own column and adding the years as an attribute.

**TRANSPOSE**


Let's tranpose the table which will group the values by sector instead of by year. The year will become the row names instead.
```{r}
# Create a new dataframe called Sector and transpose the original
Sector <- as.data.frame(t(inventory_sector))
Sector
```

**ATTRIBUTE NAMES**


Let's clean up the attribute names and remove the first row from the table since it just contains the old column names as values.
```{r}
# Clean up the sector names
colnames(Sector)[colnames(Sector) == "V1"] <- "Energy"
colnames(Sector)[colnames(Sector) == "V2"] <- "Agriculture"
colnames(Sector)[colnames(Sector) == "V3"] <- "Industrial"
colnames(Sector)[colnames(Sector) == "V4"] <- "Waste"
colnames(Sector)[colnames(Sector) == "V5"] <- "Forestry"
colnames(Sector)[colnames(Sector) == "V6"] <- "Gross"
colnames(Sector)[colnames(Sector) == "V7"] <- "Net"

# Remove first row
Sector <- Sector[-1,]
Sector
```

## Year
Create a "year" attribute in the Sector table so that it is easier for us to visualize the data later on.
```{r}
year <- rownames(Sector)
Sector$Year <- year

Sector
```

# **Predictions**
## Behavior of Greenhouse Gases
Using the collected data on air quality predict the behavior of greenhouse gases and their source. This prediction is important because greenhouse gasses are one of the main drivers of observed climate change. I would like to predict that as Sectors like Energy, Industrial and Waste emit more and more air pollutants Sectors like Agriculture and Forestry decrease.

Simply saying that as more pollutants are released into the air, the more air pollutants are absorbed by forests/land use.

## Drop NA Values
```{r}
new_sector <- as_tibble(Sector) %>% subset(!is.na(as.numeric(Energy)) & !is.na(as.numeric(Forestry)))
set.seed(385)
```

## Create Training & Test Sets
```{r}
sample_selection <- createDataPartition(as.numeric(Sector$Energy), p = 0.70, list = FALSE)
train <- new_sector[sample_selection,]
test <- new_sector[-sample_selection,]
```

## Build Model
```{r}
train_model <- lm(Sector,formula = as.numeric(Energy) ~ as.numeric(Forestry) + as.numeric(Waste) + as.numeric(Agriculture) + as.numeric(Industrial))
```

## Predictions
```{r}
predictions <- train_model %>% predict(test)

# The square of correlation - (0 - no correlation & 1 - directly correlated)
R2(predictions, as.numeric(test$Energy))

# Average errors of all our predictions
MAE(predictions, as.numeric(test$Energy))
```

# **Data Models**
## Prepare Data
Prepare the data in order for us to create a simple model. We will gather the data needed in order to graph each sector's emissions grouped by sector.
```{r}
emissions_by_sector <- gather(Sector,Sector,Emissions,"Energy":"Forestry")
# We won't need the Net or Gross, let's remove them
emissions_by_sector$Net = NULL
emissions_by_sector$Gross = NULL

# Change the type of Emissions from character to a numeric
emissions_by_sector$Emissions <- as.numeric(emissions_by_sector$Emissions)

emissions_by_sector
```

## Our Data
```{r}
emissions_model <- ggplot(emissions_by_sector, aes(x=Year,y=Emissions)) + geom_line(aes(group = Sector, color = Sector))

# Fix the x-axis so it is readable
emissions_model <- emissions_model + theme(axis.text.x = element_text(angle=45), plot.title = element_text(size = 20, face = "bold"))

# Add more detail to the labels
emissions_model <- emissions_model + labs(title = "Emissions by Inventory Sector", subtitle = "Measured in Million Metric Tons over Time (1990-2017)")

emissions_model
```

## Correlation Testing
How much two variables are related, possible outcomes can be positive, negative, or no correlation at all. Here we will focus on the **p-value**. The **p-value** describes the probability of coming about this relationship by chance.This questions if this correlation is even significant. The lower the p-value, the more likely we are observing a genuine relationship.

```{r}
# Changes the types to a numeric
Sector$Energy <- as.numeric(Sector$Energy)
Sector$Agriculture <- as.numeric(Sector$Agriculture)
Sector$Industrial <- as.numeric(Sector$Industrial)
Sector$Waste <- as.numeric(Sector$Waste)
Sector$Forestry <- as.numeric(Sector$Forestry)

# Tests the correlation between Energy
cor.test(Sector$Energy,Sector$Industrial)
cor.test(Sector$Energy,Sector$Agriculture)
cor.test(Sector$Energy,Sector$Waste)
cor.test(Sector$Energy,Sector$Forestry)
```

## Visualize the Data
Here we visualize the correlation between Energy and Forestry. I chose Forestry because it seems to have the lowest p-value **(0.02155885)** thus might be the most helpful predicter for our model. Although this has the lowest p-value, the correlation is still not very strong.
```{r}
# Create a scatter plot
forestry_vs_energy <- ggplot(Sector,aes(x=as.numeric(Sector$Forestry),y=as.numeric(Sector$Energy))) + geom_point() 

# Beautify the graph
forestry_vs_energy <- forestry_vs_energy + labs(title = "Forestry vs. Energy") + xlab("Forestry") + ylab("Energy") + theme(plot.title = element_text(size = 20, face = "bold"))

forestry_vs_energy
```

# Simple Model
In these results we should pay attention to the **coefficient**. The **coefficient** is the multiplier for our independent variable. The slope is the "Estimate". We should also pay attention the **R-squared** value.
```{r}
simple_model <- lm(Sector,formula=Energy~Forestry + Waste + Agriculture + Industrial)
summary(simple_model)
```

## Results
Our low p-value indicates that our predicters might be a meaningful addition to our model because changes in the predicter's value are related to the behavior of our response variable.

# **Parameters**
The parameters in our model is the emissions of each sector over time. This accurately shows us which sectors emit the most air pollutants. 

## Limitations
Some possible limitations of the supplemental data is that there may be more sectors that can be added to this. The sectors are vague and broad, having more detailed categories as we progress can assist us in a more detailed study of which sectors produce the most air pollutants. This also still reports on air pollutants produced in the United States, data on air pollutants produced by other countries would be helpful as well. I also realize that my predictions in the beginning may be too broad and should be narrowed down to be more effective.
