---
title: "Discovery & Data Preparation"
author: "Asheela Magwili"
output:
  html_document:
    df_print: paged
---

# **Introduction**
Researching data on air pollution is motivated by the current climate crisis. Worldwide, there are climate strikes happening due to the lack of care society has given to our planet. 

# **Data**
The current dataset included are annual emission rates recorded in 8-hour increments of toxic air pollutant chemicals like carbon dioxide, sulfur dioxide, ozone, and nitrogen oxides at the ground level.

<span style="color:red">Side Note:</span>
The necessary dataset used in the project is too large to include and therefore will need to be downloaded locally before hand. The original dataset is available for download here:https://www.kaggle.com/sogun3/uspollution. In this project, the dataset will be referred to as "uspollution.csv"

## Load Libraries
The following libraries are needed for cleaning and visualizing data in this first phase.
```{r echo=FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}
# include(library_name) installs necessary packages for a given library name
include <- function(library_name){
  if( !(library_name %in% installed.packages()) )
    install.packages(library_name) 
  library(library_name, character.only=TRUE)
}

include("tidyverse")
include("ggplot2")
```

## Load Data
Load in the dataset from the source and call it **pollution**
```{r}
pollution <- read_csv("uspollution.csv")
head(pollution)
```

## Source
This dataset's source is the Air Markets Program Database from the United States Environmental Protection Agency. Some possible limitations of this data may be that it is restricted to data collected only in the US.

## Variables

This dataset consists of twelve variables, listed below:
```{r}
cat(colnames(pollution),sep = "\n")
```

Non-obvious variables are described respectively for each pollutant below:                 
**1. Site Num: ** A unique number within the county identifying the site.                 
**2. Date Local: ** The calendar date of the average in Local Standard Time at the monitor.                 
**3. Mean: ** The average (arithmetic mean) value including all data reported in the 8-hour block.                 
**4. 1st Max Value: ** The highest value for the year.                 
**5. 1st Max Hour: ** The hour (on a 24-hour clock) when the highest value for the day (the previous field) was taken.                 
**6. AQI: ** The Air Quality Index for the day for the pollutant, if applicable.

# **Tidying Up**
## Minimize Variables
Remove some unnecessary variables that we probably won't need for this study.

1. Remove the pollutant units variable since it is redundant, all the values for each pollutant is the same throughout all the observations. It it more helpful to document what units each pollutant use rather than store this as an individual attribute.
```{r}
# Removes the variables by setting them to NULL
pollution$`NO2 Units` <- NULL
pollution$`O3 Units` <- NULL
pollution$`SO2 Units` <- NULL
pollution$`CO Units` <- NULL
head(pollution)
```

2. Remove the first column since it is just a redundant row number variable, we don't need this because it is just numbering each observation.
```{r}
# Removes the variables by setting them to NULL
pollution$X1 <- NULL
head(pollution)
```

3. Remove the State ID, County Code, and Address. We will not need it since we will be using the Site ID as our primary key.
```{r}
# Removes the variables by setting them to NULL
pollution$`State Code` <- NULL
pollution$`County Code` <- NULL
pollution$Address <- NULL
head(pollution)
```

## Variable Names
Tidy up the variable names by making them consistent - one word and no spaces

```{r}
colnames(pollution)[colnames(pollution) == "Site Num"] <- "site_id"
colnames(pollution)[colnames(pollution) == "State"] <- "state"
colnames(pollution)[colnames(pollution) == "County"] <- "county"
colnames(pollution)[colnames(pollution) == "City"] <- "city"
colnames(pollution)[colnames(pollution) == "Date Local"] <- "date"
colnames(pollution)[colnames(pollution) == "NO2 Mean"] <- "NO2_mean"
colnames(pollution)[colnames(pollution) == "NO2 1st Max Value"] <- "NO2_max_value"
colnames(pollution)[colnames(pollution) == "NO2 1st Max Hour"] <- "NO2_max_hour"
colnames(pollution)[colnames(pollution) == "NO2 AQI"] <- "NO2_aqi"
colnames(pollution)[colnames(pollution) == "O3 Mean"] <- "O3_mean"
colnames(pollution)[colnames(pollution) == "O3 1st Max Value"] <- "O3_max_value"
colnames(pollution)[colnames(pollution) == "O3 1st Max Hour"] <- "O3_max_hour"
colnames(pollution)[colnames(pollution) == "O3 AQI"] <- "O3_aqi"
colnames(pollution)[colnames(pollution) == "SO2 Mean"] <- "SO2_mean"
colnames(pollution)[colnames(pollution) == "SO2 1st Max Value"] <- "SO2_max_value"
colnames(pollution)[colnames(pollution) == "SO2 1st Max Hour"] <- "SO2_max_hour"
colnames(pollution)[colnames(pollution) == "SO2 AQI"] <- "SO2_aqi"
colnames(pollution)[colnames(pollution) == "CO Mean"] <- "CO_mean"
colnames(pollution)[colnames(pollution) == "CO 1st Max Value"] <- "CO_max_value"
colnames(pollution)[colnames(pollution) == "CO 1st Max Hour"] <- "CO_max_hour"
colnames(pollution)[colnames(pollution) == "CO AQI"] <- "CO_aqi"
head(pollution)
```

## Date
Change the type of the variable "date" from a list to a date using **as.Date()** and make sure it's format is consistent
```{r}
# Changes the type of date from list to a "date" & makes sure the format is consistently Month/Day/Year
pollution$date <- as.Date(pollution$date, "%m/%d/%y")
```

## Create Tables with Primary Keys
Create new tables for each pollutant and the location for each observation. Each table should have "site_id" as it's primary key.
```{r}
# Create Location table
Location <- tibble(site_id = pollution$site_id, state = pollution$state, county = pollution$county, city = pollution$city, date = pollution$date)
head(Location)

# Create NO2 table
NO2 <- tibble(site_id = pollution$site_id,date = pollution$date, mean = pollution$NO2_mean, max_value = pollution$NO2_max_value, max_hour = pollution$NO2_max_hour, aqi = pollution$NO2_aqi)
head(NO2)

# Create O2 table
O3 <- tibble(site_id = pollution$site_id,date = pollution$date, mean = pollution$O3_mean, max_value = pollution$O3_max_value, max_hour = pollution$O3_max_hour, aqi = pollution$O3_aqi)
head(O3)

# Create O2 table
SO2 <- tibble(site_id = pollution$site_id,date = pollution$date, mean = pollution$SO2_mean, max_value = pollution$SO2_max_value, max_hour = pollution$SO2_max_hour, aqi = pollution$SO2_aqi)
head(SO2)

# Create CO table
CO <- tibble(site_id = pollution$site_id,date = pollution$date, mean = pollution$CO_mean, max_value = pollution$CO_max_value, max_hour = pollution$CO_max_hour, aqi = pollution$CO_aqi)
head(CO)
```

# **Visualization**

## Distribution of AQI
Visualize the distribution of Air Quality Index in all four of the pollutants and observe their range.
### AQI in Nitrogen Oxides
We can visualize the air quality within nitrogen oxide by seeing how skewed the graph is. Most of the data is concentrated to the left, below 50. According to the Environmental Protection Agency (EPA), air quality with values 0 to 50 are considered good conditions. The higher the value, the higher the pollution in the air.
```{r}
# Create a density plot using geom_density()
NO2_distribution <- ggplot(data=NO2,aes(aqi)) + geom_density(kernel = "gaussian", color="blue")

# Beautify the graph
NO2_distribution <- NO2_distribution + labs(title="Distribution of Air Quality in Nitrogen Oxides",x="Air Quality Index",y="Density")
NO2_distribution
```

### AQI in Ozone
The distribution of AQI for Ozone is similarly skewed with majority of the data falling at 50 or below. We can observe that at most locations the AQI for Ozone is around 25. According to the EPA an AQI of 0 to 50 in ozone are good air conditions.
```{r}
# Create a density plot using geom_density()
O3_distribution <- ggplot(data=O3,aes(aqi)) + geom_density(kernel = "gaussian", color="blue")

# Beautify graph
O3_distribution <- O3_distribution + labs(title="Distribution of Air Quality in the Ozone",x="Air Quality Index",y="Density")
O3_distribution
```

### AQI in Sulfur Dioxides
The distribution of AQI in sulfur dioxides is low in majority of the data locations, majority of the data falls well below 25. According to the EPA an AQI of 0 to 50 in sulfur dioxides is considered good air quality.
```{r}
# Create a density plot using geom_density()
SO2_distribution <- ggplot(data=SO2,aes(aqi)) + geom_density(kernel = "gaussian", color="blue")

# Beatify graph
SO2_distribution <- SO2_distribution + labs(title="Distribution of Air Quality in Sulfur Dioxides",x="Air Quality Index",y="Density")
SO2_distribution
```

### AQI in Carbon Dioxides
The distribution of air quality of Ozone here is skewed as well, majority of the data falls below 25. According to the EPA, AQI of Ozone between 0 and 50 have no health impacts and is considered good air quality.
```{r}
# Create a density plot using geom_density()
CO_distribution <- ggplot(data=CO,aes(aqi)) + geom_density(kernel = "gaussian", color="blue")

# Beautify graph
CO_distribution <- CO_distribution + labs(title="Distribution of Air Quality in Carbon Dioxide",x="Air Quality Index",y="Density")
CO_distribution
```

## Distribution of Means
Visualizes the distribution of means among the four different pollutants. This graphically represents the average emission rate of each pollutant in the United States recorded daily in 8-hour intervals. All four graphs are heavily skewed with most data concentrated to the left.
### Means of Nitrogen Oxides
```{r}
# Create a density plot using geom_density()
NO2_distribution <- ggplot(data=NO2,aes(mean)) + geom_density(kernel = "gaussian", color="red")

# Beautify graph
NO2_distribution <- NO2_distribution + labs(title="Distribution of Mean in Nitrogen Oxides",x="Mean",y="Density")
NO2_distribution
```

### Means of Ozone
```{r}
O3_distribution <- ggplot(data=O3,aes(mean)) + geom_density(kernel = "gaussian", color="red")
O3_distribution <- O3_distribution + labs(title="Distribution of Mean in Ozone",x="Mean",y="Density")
```

### Means of Sulfur Oxides
```{r}
# Create a density plot using geom_density()
SO2_distribution <- ggplot(data=SO2,aes(mean)) + geom_density(kernel = "gaussian", color="red")

# Beautify graph
SO2_distribution <- SO2_distribution + labs(title="Distribution of Mean in Sulfur Oxides",x="Mean",y="Density")
SO2_distribution
```

### Means of Carbon Dioxides
```{r}
# Create a density plot using geom_density()
CO_distribution <- ggplot(data=CO,aes(mean)) + geom_density(kernel = "gaussian", color="red")

# Beautify Graph
CO_distribution <- CO_distribution + labs(title="Distribution of Mean in Carbon Dioxides",x="Mean",y="Density")
CO_distribution
```

# Research Questions
In the further phases of this project I hope to explore the following topics with more details

### Air Pollution's Affects on Environmental Ecosystems
I would like to explore how the behavior of air pollution over the years affect the environmental ecosystems like the rainforest, desert, oceans, etc.

### Air Pollution on Wildlife
Seeing if air pollution has correlation on wildlife behaviors.

### Air Pollution Patterns
How air pollution's behaviors change over time, as well as identify any patterns if the exist. This would further help answer the causes and solutions to air pollution.

# Conlusion
To recap this first phase, I want to mention what my intentions are with these datasets. Although it is a very interesting dataset, it is also very big and vague to really be effective. It would be interesting to see how each state performs when Carbon Emissions are measured by sectors rather than just states.